Determining Causes of Death with ChatGPT: A Case Study of Verbal Autopsy Data in Sierra Leone from 2019-2022

### 2024-01-09 manuscript/manuscript.tex
Working on methods

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. Although the survey component has been an effective alternative to collect mortality data at scale, the COD assignment component has been criticized to be expensive and difficult to reproduce due to reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. As an alternative to physician assignment, computer algorithms have recently been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \cite{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021}. Many computer algorithms have utilized the structured questionnaire portion of the surveyed data, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help algorithms perform better than using the questionnaire alone \cite{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text \cite{changSurveyEvaluationLarge2023}. Examples of these tasks include. A widely known LLM called ChatGPT developed by OpenAI. Generative Pre-trained Transformer (GPT), GPT-1 to GPT-4 \cite{wuBriefOverviewChatGPT2023}.

@@

>>>>

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. Although the survey component has been an effective alternative to collect mortality data at scale, the COD assignment component has been criticized to be expensive and difficult to reproduce due to reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. As an alternative to physician assigned CODs, computer models, such as InterVA \cite{byassIntegratedApproachProcessing2019} and InSilicoVA \cite{mccormickProbabilisticCauseofDeathAssignment2016}, have recently been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \cite{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021}. These computer models often utilized data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \cite{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023a}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for determining CODs.

@@

The performances of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for determining CODs were evaluated using 6939 deaths in 2019 to 2022 from the Health Sierra Leone (HEAL-SL) study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}. Initially, there were 11,920 deaths collected from Electronic Verbal Autopsy (EVA), where each death was randomly assigned to two different physicians that provided their diagnoses as International Classification of Diseases Revision 10 (ICD-10) codes. The codes from the two physicians were compared to determine if they agreed using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups ICD-10 codes into a set of ranges considered to be closely similar diagnoses (in agreement). If two codes for a death were not in agreement, the death enters a reconciliation phase, where the two physicians would be provided diagnoses from each other to keep their initial code, assign the other physician's code or assign a new code. If the codes still do not agree, the death enters an adjudication phase, where a third more experienced physician evaluates both physicians' diagnoses and assigns a final code. The 11,799 deaths were filtered for cases where both physicians agreed on the assigned codes (cases that were not reconciled or adjudicated).

Physicians were able to assign CODs for 11,799 of the 11,920 deaths, where 121 of these deaths could not be assigned a COD due to missing or inadequate information. The ICD-10 codes from the 11,799 deaths were then transformed into Centre for Global Health Research 10 (CGHR-10) codes that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days) age groups. Free-text open narratives were used as input for ChatGPT-3.5 and ChatGPT-4, while structured questionnaire data transformed into WHO v1.5.1 format were used as input for InterVA-5 and InSilicoVA.
